Network Data Model and BerlinMOD Benchmark

### 2010-06-23 Algebras/Network/BerlinMODNetwork/paper/PaperBerlinMODAndNetworkDataModel.tex
Related work extended.

\documentclass[a4paper]{article}
\usepackage[left=3cm, right=3cm]{geometry}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{amssymb}
\usepackage{url}
\usepackage{float}
\usepackage{longtable}
\usepackage{array}
\usepackage{algorithmic}
\usepackage[boxed]{algorithm}
\usepackage{color}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{caption,subcaption}
\pagestyle{headings}
\newcommand{\secondo}{\textsc{Secondo}}
\newcommand{\bmodb} {BerlinMOD Benchmark}
\newcommand{\op}[1]{\textbf{#1}}
\newcommand{\dt}[1]{\textsl{\underline{#1}}}
\newcommand{\true}{\textsl{TRUE}}
\newcommand{\false}{\textsl{FALSE}}
\newcommand{\secver}{2.9.x}
%opening
\title{Network Data Model and BerlinMOD Benchmark}
\author{Simone Jandt}
\date{Last Update: \today}
\begin{document}
\maketitle
\begin{abstract}
In the past, several data models for the representation of histories of
spatio-temporal data objects have been developed. We can categorize these data models
into data models for objects moving freely in the two dimensional space and data
models for network constrained moving objects. In this paper we select two representatives,
one for each data model category, which are both implemented in the \secondo{} DBMS,
and compare their capabilities with the \bmodb{}. We describe our implementation
of the used network constrained data model, and the translation
from the \bmodb{} into the network constrained data model and show
that in our experiments the network constrained data model outperforms the data
model of free movement in the two dimensional space by orders of magnitude.
\end{abstract}
\section{Introduction}
In the past, several data models for the representation of
spatio-temporal data objects have been developed. We can categorize them into
data models for objects moving freely in two dimensional space and data models
for network constrained objects. For both categories several diffferent data models
have been presented like
\cite{DataModelDataStructureGueting,RepresentingMovingObjectsGueting,STAUPelekis,HERMESMDCPelekis}
for spatio-temporal data objects moving freely in two dimensional space (DMFS) and
\cite{DynamicTransportNetworkDing,NetworkGueting,NetworkJensen,NetworkVazirgiannis}
for spatio-temporal data objects that are constrained by given networks (NCDM),
to name just a few. Objects which are restricted to use existing networks,
like cars are restricted to use road networks, can be represented as moving point
objects in both data models, whereas objects, which are not restricted by a given
network, like people, can be represented as moving point objects only in DMFS.

@@

For our benchmark experiments, presented in this paper, we chosed two data models
one for each data model category. Both data models use the same temporal representation
and are available in \secondo{} DBMS
\cite{SecondoEnvironmentDieker,SecondoPlatformPrototypingGueting}.
So we can exclude that different DBMS or temporal
representation issues bias the results of our data model comparison with the
\bmodb{}\cite{BerlinMODVLDBDuentgen}.
The DMFS we use is the data model presented in \cite{DataModelDataStructureGueting}
(SPACE for short).
And the NCDM we use is the data model presented in \cite{NetworkGueting} (NET for short).
Our implementation of NET is described first in this paper (Section \ref{sec:implNDM}
and provides some further concepts which were only sketched in \cite{NetworkGueting}.

@@

The rest of the paper is organised as follows: Related Work is presented in Section
\ref{sec:relWork}. In the sections \ref{sec:secondo} - \ref{sec:bmodb} we give
short reviews of the underlying \secondo{} DBMS (Section \ref{sec:secondo}), the two data
models (SPACE Section \ref{sec:bmodbdatamod} and NET Section \ref{sec:netdatamod})
we choose for our comparison, and the \bmodb{} (Section \ref{sec:bmodb}).
In Section \ref{sec:implNDM} we give some information about our NET implementation,
the used operations and indexes.
The translation of the \bmodb{} data and query set into the NET representation
is described in Section \ref{sec:Translation}. The resulting experimental benchmark
setup is described in Section \ref{sec:scenario} followed by the results of our
experiments in Section \ref{sec:results}. We conclude our work in Section \ref{sec:summary}.
\section{Related Work}
\label{sec:relWork}
In the past many different spatio-temporal data models have been presented. Many
of them support only discrete spatio-temporal changes
\cite{sqlstchen,Ramadrachan,Langran1,Langran2,HunterWilliamson}
or deal only with current and future positions of continuously moving objects
\cite{MOSTWolfson}. In this paper we will focus on spatio-temporal data models
for complete histories of continuously moving objects.

To the best of our knowledge only a few of these data models have been completely
implemented into different database management systems:
\begin{enumerate}
  \item HERMES \cite{HERMESMDCPelekis, HERMESPelekis} is implemented as data
cartridge for the commercial Oracle\textregistered ORDBMS \cite{Oracle}. The
used data model is the DMFS STAU \cite{STAUPelekis}.
  \item The Dynamic Transportation Network \cite{DynamicTransportNetworkDing} is
a NCDM implemented in the open source database project PostgreSQL \cite{PostgreSQL}.
  \item In the freely available \secondo{} DBMS \cite{SecondoPlatformPrototypingGueting}
two spatio-temporal data models have been implemented.
\cite{RepresentingMovingObjectsGueting,DataModelDataStructureGueting}
as DMFS and \cite{NetworkGueting} as NCDM.
\end{enumerate}

Because \secondo{} provides a data model for each of our catgories we use \secondo{}
and the provided data models. Although the data model of HERMES provides more
capabilities than SPACE.
Both data models use the basic idea of a sliced representation for moving objects
introduced by \cite{RepresentingMovingObjectsGueting,DataModelDataStructureGueting}.
The implementation of the sliced representation in SPACE supports only linear
interpolation, whereas HERMES arc interpolation is also supported. Both data
models support time varying geometries, which change their positions and / or
extent in space and time dimension. HERMES supports discret and continuous
changes, whereas SPACE only supports continuously moving objects. HERMES uses
one spatial object for different spatial geometries, whereas SPACE
uses different spatial objects for the different spatial geometries.
The same holds for operations on topological relationships. SPACE
uses different operations, whereas HERMES only provides a single operator which
distinguishes between the different topological relationships via a ``mask'' parameter.

Although HERMES provides a greater flexibility we decided to use SPACE in our
experiments, because the \secondo{} DBMS provides besides
the SPACE data model a NCDM \cite{NetworkGueting} (see Section \ref{sec:netdatamod}),
and with the \bmodb{} \cite{BMODBVLDBDuentgen} (see \ref{sec:bmodb})
to the best of our knowledge the first benchmark for complete spatio-temporal
database systems.

Other existing benchmarks \cite{COSTBenchmarkJensen,QueriesTheodoridis}

like \cite{RepresentingMovingObjectsGueting,DataModelDataStructureGueting,STAUPelekis,HERMESMDCPelekis}
and data models dealing with network constrained moving objects (NCDM) like \cite{NetworkGueting,NetworkJensen,NetworkVazirgiannis,DynamicTransportationNetworkDing}.

and \cite{NetworkGueting} as NCDM. Therefore we describe these data models
in the sections \ref{sec:bmodbdatamod} - \ref{sec:netdatamod} in more detail.

\cite{RepresentingMovingObjectsGueting,DataModelDataStructureGueting,NetworkGueting,NetworkJensen,NetworkVazirgiannis,DynamicTransportationNetworkDing},As mentioned before we can categorize spatio-temporal data models into
data models dealing with current and future positions of moving objects,
data models dealing with complete histories of moving objects, and data models
dealing with past, current and future positions of moving objects. We will focus
on data models dealing with complete histories of moving objects, like \cite{STAUPelekis,HERMESMDCPelekis,RepresentingMovingObjectsGueting,DataModelDataStructureGueting,NetworkGueting,NetworkJensen,NetworkVazirgiannis,DynamicTransportationNetworkDing}, to name just a few.

DMFS\cite{STAUPelekis,HERMESPelekis,RepresentingMovingObjectsGueting,DataModelDataStructureGueting,sqlstchen}

NCDM\cite{NetworkGueting,NetworkJensen,NetworkVazirgiannis,DynamicTransportationNetworkDing}.

Also, there are some more
benchmarks \cite{COSTBenchmarkJensen,QueriesTheodoridis} and database systems for
spatio-temporal data types \cite{SecondoPlatformPrototypingGueting,HERMESPelekis}. But to the best of our
knowledge they do not provide a combination of different implemented
and supported data models together with an existing benchmark feasible for moving
objects in space and network constrained moving objects like the current \secondo{} DBMS.
So it is natural for us to use the \secondo{} DBMS in combination with the provided
data models and the \bmodb{} to compare the capabilities of the two data models.

In the next sections we give short reviews of the \secondo{} database system
(Section \ref{sec:secondo}), the both data models (Section \ref{sec:bmodbdatamod},
Section \ref{sec:netdatamod}) we used in our experiments, and the \bmodb{}
(Section \ref{sec:bmodb}).
\section{\secondo{} DBMS}
\label{sec:secondo}
The extensible \secondo{} DBMS presented in \cite{SecondoEnvironmentDieker,SecondoPlatformPrototypingGueting} provides a
platform for implementing various kinds of data models. It provides a clean
interface between the data model independent system frame and the content of the
single data models. Hence \secondo{} can be easily extended by the user
implementing algebra modules to introduce new data types and operations on
these data types. The user may define additional viewers for the graphical user
interface or write additional optimization rules or cost functions to extend the
the optimizer. Since \secondo{} version 2.9 the users may publish their extensions
as a \secondo{} plugin such that other users can use this plugin to extend their own
\secondo{} system. They may use the newly provided functionalities or repeat the
published experiments. \secondo{} is freely available on the web \cite{secondoweb}.
It comes with a number of already implemented spatial and spatio-temporal data types
and operations including the DMFS (Section \ref{sec:bmodbdatamod}) and the NDM
(Section \ref{sec:netdatamod}) used in this paper. Furthermore, the \bmodb{}
described in Section \ref{sec:bmodb} has been developed in the \secondo{} DBMS.
For our experiments we used the \secondo{} version \secver{}.
\section{Data Model of Free Movement in Two Dimensional Space(SPACE)}
\label{sec:bmodbdatamod}
The data model used by the \bmodb{} is the DMFS presented in \cite{STDTModelingQueryingMOinDErwig,DataModelDataStructureGueting,RepresentingMovingObjectsGueting}.
Spatial positions are assumed to be located in a two dimensional space.

A single spatial
position is represented by the data type \dt{point}. A \dt{point} consists of a pair of
\dt{real} values interpreted as x,y-coordinates in the assumed two dimensional plane.

Streets are represented by \dt{line} values. A \dt{line} value consists of a set
of \dt{HalfSegment}s representing the geometry of the line in the two dimensional space.
Each \dt{HalfSegment} consists of two \dt{point} values and a Boolean flag telling us
if the left or the right point is the dominating point of the \dt{HalfSegment}.

Regions are represented by the data type \dt{region}. A \dt{region} consists of
a set of \dt{HalfSegment}s defining the outer border of the region in
the two dimensional space. If a region contains wholes the inner border is also
formed by the \dt{HalfSegment}s.

In \secondo{} all these spatial data types and many standard data types can be
''lifted'' to become time dependent \dt{moving} values. For all data types \dt{$\alpha$}
the constructor \dt{moving} creates a new data type \dt{moving}(\dt{$\alpha$})
(short form \dt{m$\alpha$}).

@@

Some other data types of \secondo{} which are used in the \bmodb{} are shown in
Table \ref{tab:bmodbdatatypes}.
\begin{table}[H]
\begin{center}
\begin{scriptsize}
\begin{tabularx}{1.0\textwidth}{|l|X|}
\hline
\textbf{Data Type} & \textbf{Description} \\
\hline
\dt{bool} & Usual boolean data type\\
\hline
\dt{int} & Usual integer number\\
\hline
\dt{real} & Usual real number\\
\hline
\dt{instant} & A point in time\\
\hline
\dt{periods} & A set of disjoint and not connected time interval\\
\hline
\dt{mbool} & A time dependent boolean value, which is constant \true{} or \false{}
within each \dt{ubool} \\
\hline
\dt{mreal} & Time dependent real number. Each unit is defined by a function
of time representing the \dt{real} value at each time instant.\\
\hline
\end{tabularx}
\end{scriptsize}
\caption{Other Data Types of \bmodb{}}
\label{tab:bmodbdatatypes}
\end{center}
\end{table}
\section{Network Data Model(NET)}
\label{sec:netdatamod}
The central idea of the NDM presented in \cite{NetworkGueting} is that
every movement is constrained by a given network and every position can be described
relative to this network. The data type \dt{network} is the central
data type in the NDM. All other data types of the NDM
are related to a given \dt{network} object by the unique network identifier that
is part of each \dt{network} object.

@@

The spatio-temporal minimum bounding box was introduced as an parameter to the
\dt{mgpoint} because the computation of this value is very expensive in the
NDM. Although each unit of an \dt{mgpoint} stays on the same route
at the same speed it may follow different spatial directions. For example, a route may
lead uphill in serpentine. A spatial bounding box only computed from the spatial
start and end position may not enclose all spatial positions of the car within
the unit. Therefore we always have to examine the spatial dimensions of the
\dt{RouteInterval} passed within a unit to compute the units bounding box. This
needs an access to the route curve in the $routes$ relation of the corresponding
\dt{network} object. If $r$ is the number of routes
of the network and $h$ the number of \dt{HalfSegment}s belonging to the \dt{RouteInterval}
passed in a unit, we need O($h + \log r$) time to
compute the bounding box for a single unit. The bounding box of the \dt{mgpoint}
is the union of the bounding boxes of its $m$ units. So the computation of the
spatio-temporal respectively spatial bounding box of an \dt{mgpoint} needs
O($H + m \log r)$) time, with $H = \sum_{i=1}^m {h_i}$ if $h_i$ is the number of the
$half$ $segments$ for unit $i$, $1 \leq i \leq m$ of the \dt{mgpoint}. This very expensive computation is only done
on demand or if we can get the bounding box for free. For example we can copy
the bounding box of an \dt{mpoint} at the translation time into an \dt{mgpoint}
without big computational costs. The bounding box attribute is not maintained. If
the \dt{mgpoint} value changes, the bounding box attribute is set to be undefined
until recomputing is necessary.
\section{BerlinMOD Benchmark}
\label{sec:bmodb}
The \bmodb{} was presented in \cite{BerlinMODVLDB} and the
provided scripts for the data generator are implemented as \secondo{} DBMS operations.
The \bmodb{} is available on the web \cite{berlinmodweb} and provides a well defined
data-set and queries for the experimental evaluation of the capabilities of
spatial and spatio-temporal database systems dealing with histories of moving
objects. The \bmodb{} emphasises the development of complete systems
and simplifies experimental repeatability pointing out the capabilities and the
weaknesses of the benchmarked systems.

>>>>

\documentclass[a4paper]{article}
\usepackage[left=3cm, right=3cm]{geometry}
\usepackage[utf8x]{inputenc}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{amssymb}
\usepackage{url}
\usepackage{float}
\usepackage{longtable}
\usepackage{array}
\usepackage{algorithmic}
\usepackage[boxed]{algorithm}
\usepackage{color}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{caption,subcaption}
\pagestyle{headings}
\newcommand{\secondo}{\textsc{Secondo}}
\newcommand{\bmodb} {BerlinMOD Benchmark}
\newcommand{\op}[1]{\textbf{#1}}
\newcommand{\dt}[1]{\textsl{\underline{#1}}}
\newcommand{\true}{\textsl{TRUE}}
\newcommand{\false}{\textsl{FALSE}}
\newcommand{\secver}{2.9.x}
%opening
\title{Network Data Model and BerlinMOD Benchmark}
\author{Simone Jandt}
\date{Last Update: \today}
\begin{document}
\maketitle
\begin{abstract}
In the past, several data models for the representation of histories of
spatio-temporal data objects have been developed. We can categorize these data models
into data models for objects moving freely in the two dimensional space and data
models for network constrained moving objects. In this paper we select two representatives,
one for each data model category, which are both implemented in the \secondo{} DBMS,
and compare their capabilities with the \bmodb{}. We describe our implementation
of the used network constrained data model, and the translation
from the \bmodb{} into the network constrained data model and show
that in our experiments the network constrained data model outperforms the data
model of free movement in the two dimensional space by orders of magnitude.
\end{abstract}
\section{Introduction}
In the past, several data models for the representation of
spatio-temporal data objects have been developed. We can categorize them into
data models for objects moving freely in two dimensional space and data models
for network constrained objects. For both categories several diffferent data models
have been presented like
\cite{DataModelDataStructureGueting,RepresentingMovingObjectsGueting,STAUPelekis,HERMESMDCPelekis}
for spatio-temporal data objects moving freely in two dimensional space (DMFS) and
\cite{DynamicTransportNetworkDing,NetworkGueting,NetworkJensen,NetworkVazirgiannis}
for spatio-temporal data objects that are constrained by given networks (NCDM),
to name just a few. Objects which are restricted to use existing networks,
like cars are restricted to use road networks, can be represented as moving point
objects in both data models, whereas objects, which are not restricted by a given
network, like people, can be represented as moving point objects only in DMFS.

@@

For our benchmark experiments, presented in this paper, we chosed two data models
one for each data model category. Both data models use the same temporal representation
and are available in \secondo{} DBMS
\cite{SecondoEnvironmentDieker,SecondoPlatformPrototypingGueting}.
So we can exclude that different DBMS or temporal
representation issues bias the results of our data model comparison with the
\bmodb{}\cite{BerlinMODVLDBDuentgen}.
The DMFS we use is the data model presented in
\cite{RepresentingMovingObjectsGueting,DataModelDataStructureGueting}
(SPACE for short).
And the NCDM we use is the data model presented in \cite{NetworkGueting} (NET for short).
Our implementation of NET is described first in this paper (Section \ref{sec:implNDM}
and provides some further concepts which were only sketched in \cite{NetworkGueting}.

@@

The rest of the paper is organised as follows: Related Work is presented in Section
\ref{sec:relWork}, including short reviews of the underlying \secondo{} DBMS
(Section \ref{sec:secondo}), the two data models (SPACE Section \ref{sec:bmodbdatamod}
and NET Section \ref{sec:netdatamod}) we chosed for our comparison, and the
\bmodb{} (Section \ref{sec:bmodb}).
In Section \ref{sec:implNDM} we give some informations about our implementation
of NET, the used operations and indexes.
The translation of the \bmodb{} data and query set into the NET representation
is described in Section \ref{sec:Translation}. The resulting experimental benchmark
setup is described in Section \ref{sec:scenario} followed by the results of our
experiments in Section \ref{sec:results}. We conclude our work in Section \ref{sec:summary}.
\section{Related Work}
\label{sec:relWork}
In the past many different spatio-temporal data models have been presented. Many
of them support only discrete spatio-temporal changes
\cite{sqlstchen,HunterWilliamson,Langran2,Langran1,Ramadrachan}
or deal only with current and future positions of continuously moving objects
\cite{MOSTWolfson}. Reviews of several different spatio-temporal data models are
given in \cite{ReviewSTDMPelekis}. In this paper we will focus on spatio-temporal
data models for complete histories of continuously moving objects. These can be
categorized into data models for objects moving freely in two dimensional space
(DMFS) and data models for network constrained moving objects (NCDM).

We will start with a few DMFS. \cite{MOLRaQSu} proposes the basic ideas of an
DMFS based on differential geometry and the linear constraints approach of modeling
spatial and spatio-temporal data. Spatial and spatio-temporal objects
are defined by linear constraints, and the queries use formulas from differential
geometry. The paper proposes only an incomplete abstract model without any usability.

\cite{STAUPelekis} and \cite{RepresentingMovingObjectsGueting,DataModelDataStructureGueting}
use both the idea of time sliced representation of moving objects first presented
in \cite{RepresentingMovingObjectsGueting}. \cite{DataModelDataStructureGueting}
supports only linear interpolation of movement, whereas \cite{STAUPelekis} also
supports arc interpolation. \cite{STAUPelekis} uses only one spatial
object containing all spatial geometries and one moving object for the
representation of the different moving object data types, whereas
\cite{RepresentingMovingObjectsGueting,DataModelDataStructureGueting} uses different
spatial and moving objects for the representation of the different
data types. Therefore \cite{STAUPelekis} provides only a single operator that
distinguishes between the different topological relationships via a parameter,
whereas \cite{RepresentingMovingObjectsGueting,DataModelDataStructureGueting}
uses different operations to estimate topological relationships. Alltogether
\cite{STAUPelekis} offers a more flexible object oriented design than
\cite{RepresentingMovingObjectsGueting,DataModelDataStructureGueting}.

The spatio-temporal framework proposed by
\cite{DataModelDataStructureGueting,RepresentingMovingObjectsGueting} is extended
by \cite{NetworkGueting} to be also usable in network constrained environments.
Similar to the most other NCDM in this section \cite{NetworkGueting} depends on
a fixed network representation. That means that the old network informations are lost when a
update of the network data structure occurs. Only \cite{DynamicTransportNetworkDing}
can handle such updates without loss of information.

\cite{NetworkVazirgiannis} proposes an edge based NCDM. With an relation holding
the edges and their attributes. The edge relation represents a undirected graph.
Moving objects are assumed to drive always on the path with the lowest cost
(distance or travel-time) from the source to the target point from a given
starting time. The trajectory is computed by this assumptation using the length
and speed attributes of the graph edges within the shortest / fastest path computation.
These assumptions will not always hold in real world environments. Such that the
moving object representation does not suffice real world requirements.

\cite{NetworkJensen} uses also an edge based network representation. The paper
proposes a combination of an 2D geometrical edge representation with an directed
graph representation of the network. Where as the 2D geometrical representation
handles the spatial information the connectivity information is mostly embedded
in the directed graph representation. The both representations are connected by
transition policies. Moving objects are represented by sets of 5 tuples. Each
5 tuple contains an edge identifier, the position of moving point on the edge
in terms of weight and length, the speed and direction of the movement, and
the time instant of this information.

Different from all other NCDM in this section \cite{DynamicTransportNetworkDing}
proposes a dynamic transportation network model. This means that the attributes
of the network parts are modelled by time depending dynamic attributes, such that
changes in the network environment can be handled without loss of information.
Similar to \cite{NetworkGueting} the idea of the network model bases
on routes and junctions. The authors propose a two-layer network representation
data structure, which combines the advantages of dynamic edge based
\cite{DynNetworkModEdgeGueting} and route based \cite{DynNetworkModRouteGueting}
NCDM approach. The route based environment reduces the update intervals and
used storage space for moving objects, whereas the edge-based envirionment supports
a more detailed view on the traffic conditions of the different edges belonging
to the same route. Moving objects are represented by a set of pairs of
motion vectors together with a boolean flag telling if the motion vector contains
actual or historical information. Each motion vector consists, in parts
similar to \cite{NetworkJensen}, of a time stamp, a network position, and a speed vector.

To the best of our knowledge only a few of these data models have been completely
implemented into different database management systems:
\begin{enumerate}
  \item \cite{STAUPelekis} is implemented as data cartridge \cite{HERMESMDCPelekis,
HERMESPelekis} for the commercial Oracle\textregistered ORDBMS \cite{ORACLE}.
  \item \cite{DynamicTransportNetworkDing} is implemented as extension of the
open source database project PostgreSQL \cite{PostgreSQL}.
  \item \cite{RepresentingMovingObjectsGueting,DataModelDataStructureGueting}
and \cite{NetworkGueting} are implemented in the freely available extensible
\secondo{} DBMS \cite{SecondoPlatformPrototypingGueting}.
\end{enumerate}

Although \cite{STAUPelekis} for the DMFS and \cite{DynamicTransportNetworkDing} for
the NCDM provide greater flexibility we decided to use
\cite{RepresentingMovingObjectsGueting,DataModelDataStructureGueting}
and \cite{NetworkGueting} in our experiments, because both are available in the
freely available \secondo{} DBMS. And not at least \secondo{} is the DBMS with
which the \bmodb{} \cite{BerlinMODVLDBDuentgen} was developed.

Other benchmarks for spatio-temporal databases provide only well defined
query sets \cite{QueriesTheordoridis} and a database description without data.
Or they come with well defined data generation, workload sets and experiments but
only to evaluate the capabilities of indexes for current and near future positions
\cite{COSTBenchmarkJensen}. Or they focus on time-evolving regional data and associated
index methods like \cite{BenchmarkTzouramanis}.

The \bmodb{} is to the best of our knowledge the only benchmark testing the capabilities
of complete spatio-temporal database systems. Coming with well defined data and
query sets feasible for moving objects in space and network constrained moving objects.

In the next sequel we give short reviews of the \secondo{} database system
(Section \ref{sec:secondo}), the both data models (Section \ref{sec:bmodbdatamod},
Section \ref{sec:netdatamod}) we used in our experiments, and the \bmodb{}
(Section \ref{sec:bmodb}).
\subsection{\secondo{} DBMS}
\label{sec:secondo}
The extensible \secondo{} DBMS presented in
\cite{SecondoEnvironmentDieker,SecondoPlatformPrototypingGueting} provides a
platform for implementing various kinds of data models. It provides a clean
interface between the data model independent system frame and the content of the
single data models. Hence \secondo{} can be easily extended by the user
implementing algebra modules to introduce new data types and operations on
these data types. The user may define additional viewers for the graphical user
interface or write additional optimization rules or cost functions to extend the
the optimizer. Since \secondo{} version 2.9 the users may publish their extensions
as a \secondo{} plugin such that other users can use these plugins to extend their own
\secondo{} system. They may use the newly provided functionalities or repeat the
published experiments. \secondo{} is freely available on the web \cite{secondoweb}.
It comes with a number of already implemented spatial and spatio-temporal data types
and operations including the DMFS (Section \ref{sec:bmodbdatamod}) and the NCDM
(Section \ref{sec:netdatamod}) used in this paper. Furthermore, the \bmodb{}
described in Section \ref{sec:bmodb} has been developed in the \secondo{} DBMS.
For our experiments we used the \secondo{} version \secver{}.
\subsection{Data Model of Free Movement in Two Dimensional Space}
\label{sec:bmodbdatamod}
\cite{STDTModelingQueryingMOinDErwig} presents the basic idea of the new
spatio-temporal data model that we will use as DMFS in our experiments. Besides
the general idea it discusses related work and the advantages of having abstract
and discrete data models. Abstract data models allow us to make definitions
in terms of infinite sets, without worrying whether finite representations exist.
They are useful as conceptual model, but they cannot be implemented, because computers
can only use finite sets. Data structures and algorithms have to work with finite
representations of the data types. Therefore we also need a discrete algebra
representation. The abstract definition of our data model has been published in
\cite{RepresentingMovingObjectsGueting}, whereas an associated discrete data
model was presented in \cite{DataModelDataStructureGueting}.

The type system in \cite{RepresentingMovingObjectsGueting} is defined using the
techniques presented in \cite{SecondOrderSignatureGueting}. The basic idea of
\cite{RepresentingMovingObjectsGueting} is to define type constructors
that create new data types if they are applied to an data type of a given set of
basic data types.
Basic data types are the standard data types like integer, real, string and Boolean (BASE);
the spatial data types point, points, line, and region (SPATIAL);
and the temporal type instant (TIME).

The data types of BASE are defined as usual. The SPATIAL data types are defined
by different carrier sets in the abstract and the disrete level.

In the abstract data model the carrier set of a point is $\mathbb{R}^2 \cup \{ \perp \}$.
In the discret data model the carrier set of a point is $\{ real \times real \} \cup \{ \perp \}$,
whereas the two $real$ values are interpreted as x,y-coordinates of a point in the
two dimensional plane.

In the abstract data model line and region are subsets of the set of point that
fullfill several conditions. In the discret data model a \dt{line} is a linear
approximation of the curve of the line by a finite set of line segments in the plane.
Regions are represented by a set of segments representing a linear approximation
of the region borders by polygons in the discrete data model.

The two most important type constructors are $moving$ and $range$. $Moving$
maps each type $\alpha$ from BASE and SPATIAL into an time dependend spatio-temporal
moving data type $moving$($\alpha$) (TEMPORAL), whereas $range$ is applicable to
BASE and TEMPORAL type $\beta$ and produces range data types $range$($\beta$).

In the abstract data model a time instant is a point in time or undefined, whereas
the carrier set for instant is $\mathbb{R} \cup \{ \perp \}$.
In \secondo{} all these spatial data types and many standard data types can be
''lifted'' to become time dependent \dt{moving} values. For all data types \dt{$\alpha$}
the constructor \dt{moving} creates a new data type \dt{moving}(\dt{$\alpha$})
(short form \dt{m$\alpha$}).

@@

Some other data types of \secondo{} which are used in the \bmodb{} are shown in
Table \ref{tab:bmodbdatatypes}.
\begin{table}[H]
\begin{center}
\begin{scriptsize}
\begin{tabularx}{1.0\textwidth}{|l|X|}
\hline
\textbf{Data Type} & \textbf{Description} \\
\hline
\dt{bool} & Usual boolean data type\\
\hline
\dt{int} & Usual integer number\\
\hline
\dt{real} & Usual real number\\
\hline
\dt{instant} & A point in time\\
\hline
\dt{periods} & A set of disjoint and not connected time interval\\
\hline
\dt{mbool} & A time dependent boolean value, which is constant \true{} or \false{}
within each \dt{ubool} \\
\hline
\dt{mreal} & Time dependent real number. Each unit is defined by a function
of time representing the \dt{real} value at each time instant.\\
\hline
\end{tabularx}
\end{scriptsize}
\caption{Other Data Types of \bmodb{}}
\label{tab:bmodbdatatypes}
\end{center}
\end{table}
\subsection{Network Data Model(NET)}
\label{sec:netdatamod}
The central idea of the NDM presented in \cite{NetworkGueting} is that
every movement is constrained by a given network and every position can be described
relative to this network. The data type \dt{network} is the central
data type in the NDM. All other data types of the NDM
are related to a given \dt{network} object by the unique network identifier that
is part of each \dt{network} object.

@@

The spatio-temporal minimum bounding box was introduced as an parameter to the
\dt{mgpoint} because the computation of this value is very expensive in the
NDM. Although each unit of an \dt{mgpoint} stays on the same route
at the same speed it may follow different spatial directions. For example, a route may
lead uphill in serpentine. A spatial bounding box only computed from the spatial
start and end position may not enclose all spatial positions of the car within
the unit. Therefore we always have to examine the spatial dimensions of the
\dt{RouteInterval} passed within a unit to compute the units bounding box. This
needs an access to the route curve in the $routes$ relation of the corresponding
\dt{network} object. If $r$ is the number of routes
of the network and $h$ the number of \dt{HalfSegment}s belonging to the \dt{RouteInterval}
passed in a unit, we need O($h + \log r$) time to
compute the bounding box for a single unit. The bounding box of the \dt{mgpoint}
is the union of the bounding boxes of its $m$ units. So the computation of the
spatio-temporal respectively spatial bounding box of an \dt{mgpoint} needs
O($H + m \log r)$) time, with $H = \sum_{i=1}^m {h_i}$ if $h_i$ is the number of the
$half$ $segments$ for unit $i$, $1 \leq i \leq m$ of the \dt{mgpoint}. This very expensive computation is only done
on demand or if we can get the bounding box for free. For example we can copy
the bounding box of an \dt{mpoint} at the translation time into an \dt{mgpoint}
without big computational costs. The bounding box attribute is not maintained. If
the \dt{mgpoint} value changes, the bounding box attribute is set to be undefined
until recomputing is necessary.
\subsection{BerlinMOD Benchmark}
\label{sec:bmodb}
The \bmodb{} was presented in \cite{BerlinMODVLDB} and the
provided scripts for the data generator are implemented as \secondo{} DBMS operations.
The \bmodb{} is available on the web \cite{berlinmodweb} and provides a well defined
data-set and queries for the experimental evaluation of the capabilities of
spatial and spatio-temporal database systems dealing with histories of moving
objects. The \bmodb{} emphasises the development of complete systems
and simplifies experimental repeatability pointing out the capabilities and the
weaknesses of the benchmarked systems.
